---
title: "Chocolate and nobel prize"
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

From *Chocolate and the Nobel Prize – a true story?*

http://gforge.se/2012/12/chocolate-and-nobel-prize/

First we get nobel prizes per country by reading the table using readHTMLTable() from the XML package. After that we do some cleaning to the dataset.

```{r, message=FALSE, warning=FALSE, comment=FALSE}
rm(list=ls())
library(XML)
library(httr)
url <- "http://en.wikipedia.org/wiki/List_of_countries_by_Nobel_laureates_per_capita"
theurl <- htmlParse(rawToChar(GET(url)$content))
tables <- readHTMLTable(theurl)
# delete  Faroe Islands
tables <- tables[[2]][-c(1,31),]

nobel_prizes <- tables
# Clean column names
colnames(nobel_prizes) <- 
  gsub(" ", "_", 
       gsub("(/|\\[[0-9]+\\])", "", 
            gsub("\n", " ", colnames(nobel_prizes))
        )
    )
 
# Delete those that aren't countries and thus lack rank
nobel_prizes$Rank <- as.numeric(as.character(nobel_prizes$Rank))
nobel_prizes <- subset(nobel_prizes, is.na(Rank) == FALSE)

# Create Country
nobel_prizes$Country = nobel_prizes$Entity

# Clean the country names
nobel_prizes$Country <- 
  gsub("[^a-zA-Z ]", "", nobel_prizes$Entity)
 
# Clean the loriates variable
nobel_prizes$Laureates_10_million <- 
  as.numeric(as.character(nobel_prizes$Laureates_10_million))

```

The next part is slightly trickier since we need to translate german country names to match the nobel prize data.

```{r, message=FALSE, warning=FALSE, comment=FALSE}
# Translation from German to English
url <- "http://german.about.com/library/blnation_index.htm"
theurl <- htmlParse(rawToChar(GET(url)$content))
tables <- readHTMLTable(theurl)
 
translate_german <- tables[[1]]
translate_german <- translate_german[3:NROW(translate_german), 1:2]
colnames(translate_german) <- c("English", "German")
translate_german$German <- 
    gsub("([ ]+(f|pl)\\.$|\\([[:alnum:] -]+\\))", "", translate_german$German)

# Get the consumption from a German list
url <- "http://www.theobroma-cacao.de/wissen/wirtschaft/international/konsum"
theurl <- htmlParse(rawToChar(GET(url)$content))

tables <- readHTMLTable(theurl)

german_chocolate_data <- tables[[1]][2:NROW(tables[[1]]), ]
names(german_chocolate_data) <- c("Country", "Chocolate_consumption")
german_chocolate_data$Country <- 
    gsub("([ ]+f\\.$|\\([[:alnum:] -]+\\))", "", german_chocolate_data$Country)

library(sqldf)

sql <- paste0("SELECT gc.*, tg.English as Country_en",
              " FROM german_chocolate_data AS gc", 
              " LEFT JOIN translate_german AS tg", 
              " ON gc.Country = tg.German",
              " OR gc.Country = tg.English")
german_chocolate_data <- sqldf(sql)

german_chocolate_data$Country <- 
    ifelse(is.na(german_chocolate_data$Country_en), 
    german_chocolate_data$Country, 
    german_chocolate_data$Country_en)
    
german_chocolate_data$Country_en <- NULL
german_chocolate_data$Chocolate_consumption_tr <- NA
for (i in 1:NROW(german_chocolate_data)) {
    number <- as.character(german_chocolate_data$Chocolate_consumption[i])
    if (length(number) > 0) {
        m <- regexpr("^([0-9]+,[0-9]+)", number)
        if (m > 0) {
            german_chocolate_data$Chocolate_consumption_tr[i] <- 
                as.numeric(
                    sub(",", ".", regmatches(number, m))
                )
        } else {
            m <- regexpr("\\(([0-9]+,[0-9]+)", number)

            if (m > 0) 
                german_chocolate_data$Chocolate_consumption_tr[i] <- 
                    as.numeric(
                        sub("\\(", "", 
                            sub(",", ".", regmatches(number, m))
                        )
                    )
        }
    }
}

sql <- paste0("SELECT np.*, gp.Chocolate_consumption_tr AS choc",
              " FROM nobel_prizes AS np", 
              " LEFT JOIN german_chocolate_data AS gp", 
              " ON gp.Country = np.Country")
nobel_prizes <- sqldf(sql)
```

Chocalate data  for Switzerland

```{r, message=FALSE, warning=FALSE}
nobel_prizes$choc[nobel_prizes$Country == "Switzerland"] <- 11.9
```

This leaves us with 18 countries that have chocolate data. 

```{r, message=FALSE, warning=FALSE}
sum(complete.cases(nobel_prizes))
nobel_prizes_cc = nobel_prizes[complete.cases(nobel_prizes),]
nobel_prizes_cc
```

Lets plot Countries' Annual Per Capita Chocolate Consumption and the Number of Nobel Laureates per 10 Million Population (in the log scale)

```{r, message=FALSE, warning=FALSE}
cc = nobel_prizes_cc$choc
paesi = nobel_prizes_cc$Country
pop_temp = as.character(nobel_prizes_cc$`Population_(2017)[note_1]`)
pop = as.numeric(gsub(",", "", pop_temp))
np = 10^7*as.numeric(nobel_prizes_cc$`Nobel_laureates_(2017)`)/pop
lnp = log(np)
cor(lnp,cc)
plot(cc, np, log="y")
text(cc, np, labels=paesi )
```

Without controlling for other factors:

```{r, message=FALSE, warning=FALSE}
plot(cc, lnp)
abline(lm(lnp ~ cc), lwd=2, col=2)
summary(lm(lnp ~ cc))
```

What else might explain this relationship? 
How do we identify confounders?
How do we control for them? 

Best case: randomize treatment. In a randomized experiment, there
should be no confounding variables.

In many social science settings, RCT is
impossible: subjects (e.g. countries, individuals)
choose their own treatment.


The Nordic countries rank at the top of the graph on the right and they are known to rank high on per capita income. 

Controlling for Nordic countries and GDP?

```{r}
library("countrycode")
countryc = countrycode(paesi, origin="country.name", destination="wb")
library("wbstats")
new_wb_cache <- wbcache()
wbsearch("gdp.*capita.*US\\$", cache = new_wb_cache)
indicator <- wb(indicator = c("NY.GDP.PCAP.KD"), startdate=2010, enddate=2010, country = countryc) 
gdp <- indicator$value[match(paesi, indicator$country)]
nc = paesi %in% c("Sweden","Denmark","Norway","Canada", "Finland", "Ireland","Austria","Switzerland", "Germany", "Belgium")*1
plot(cc,lm(lnp ~ gdp + nc)$resid)
abline(lm( lm(lnp ~ gdp + nc)$resid ~ cc ), lwd=2,col=2 )
summary(lm(lnp ~ cc + gdp + nc))
```


