---
title: "Algorithms and Inference"
subtitle: "Statistical Learning"
author: ""
output: pdf_document
---

From Efron and Hastie (2016), chapter 1

## Algorithms and Inference


Statistics is the science of *learning from experience*, particularly experience that arrives a little bit at a time

* the successes and failures of a new
experimental drug

* the uncertain measurements of an asteroid’s path toward
Earth

* etc.

There are two main aspects of statistical analysis:

* the *algorithmic* aspect
* the *inferential* aspect

The distinction begins with the
most basic, and most popular, statistical method, averaging.

Suppose we have observed $y_1,\ldots,y_n$, realizations of the random variables $Y_1,\ldots,Y_n$ i.i.d. $Y$, applying to some phenomenon of
interest, where we choose that the *parameter of interest* is $\mathbb{E}(Y)$.

Averaging is the \emph{algorithm}
$$\bar{y} = \frac{1}{n}\sum_{i=1}^{n}y_i$$
How accurate is that number?

The standard error provides an inference on the algorithm's accuracy 
$$\widehat{\mathrm{se}} = \sqrt{\frac{1}{n} \frac{\sum_{i=1}^{n}(y_i - \bar{y})^2}{n-1}}$$

It is a surprising, and crucial, aspect
of statistical theory that the same data that supplies an estimate can also
assess its accuracy.

Of course, $\widehat{\mathrm{se}}$ is itself an algorithm, which could be (and is) subject
to further inferential analysis concerning its accuracy. 

The point is that
the algorithm comes *first* and the inference follows at a *second* level of
statistical consideration. 

In practice this means that algorithmic invention
is a more free-wheeling and adventurous enterprise, with inference playing
catch-up as it strives to assess the accuracy, good or bad, of some hot new
algorithmic methodology.

\newpage

## Basic Gaussian model

In the basic Gaussian model with $Y_1,\ldots,Y_n$ i.i.d. $Y \sim \mathcal{N}(\mu, \sigma^2)$, where 

* the unknown parameter $\mu$ is the *parameter of interest*, and we wish to assess the relation of the data to that value 

* $\sigma^2$ is the *nuisance parameter*, which can be either known or unknown.

One well-known  *point estimator*  for $\mu$ is 

$$\hat{\mu} = \frac{1}{n}\sum_{i=1}^{n}Y_i \sim N(\mu, \sigma^2/n)$$
with standard error 

$$\mathrm{se}(\hat{\mu}) =  \sqrt{\mathrm{Var}(\hat{\mu})} = \sigma \sqrt{ 1 /n}$$

If $\sigma^2$ is unknown, we use the estimator of the standard error

$$\widehat{\mathrm{se}}(\hat{\mu}) = \hat{\sigma} \sqrt{ 1 /n}$$
where the estimator for $\sigma^2$ is

$$\displaystyle \hat{\sigma}^2 = \frac{\sum_{i=1}^{n}(Y_i - \hat{\mu})^2}{n-1} \sim \sigma^2 \chi^2_{n-1} /(n-1)$$

which is independent from $\hat{\mu}$. 

A *pivot* for inference about $\mu$ is

$$ T= \displaystyle \frac{ \hat{\mu}  - \mu}{\sigma \sqrt{1/n}} \cdot  \frac{\sigma }{\hat{\sigma}} \sim \frac{N(0,1)}{\sqrt{\chi^2_{n-1} /(n-1)}}\sim \mathcal{T}_{n-1}$$
with $\mathrm{Pr}( - t^{1-\alpha/2}_{n-1} \leq T \leq t^{1-\alpha/2}_{n-1} ) = 1-\alpha$ where $t^{1-\alpha/2}_{n-1}$ is the $1-\alpha/2$ quantile of the Student $\mathcal{T}$ distribution with $n-1$ degrees of freedom. 

A $1-\alpha$ *confidence interval* for $\mu$ is 

\begin{eqnarray}\label{ci}
[\underline{\mu},\overline{\mu}]  = \hat{\mu} \pm t^{1-\alpha/2}_{n-1} \cdot \hat{\mathrm{se}}(\bar{Y}) 
\end{eqnarray}

The *coverage* probability of the confidence interval in (\ref{ci}) is then
$$\mathrm{Pr}([\underline{\mu},\overline{\mu}]  \ni \mu ) = 1-\alpha$$

The following simulation corroborates the coverage property of the confidence interval:   

```{r}
sim = function(alpha=0.05, n=5, mu=0, sigma=1){
  y = rnorm(n, mean=mu, sd=sigma)
  hatmu = mean(y)
  hatse = sqrt( var(y) / n )
  k = qt(alpha/2, df = n-1, lower.tail = F)
  ci = hatmu + c(-1,1) * k * hatse
  cover = (mu >= ci[1] & mu <= ci[2])
  return(cover)
}
set.seed(123)
mean( replicate(100, sim(alpha=0.05) ))
```

Suppose now that the target of inference is a future realization $y$ of $Y$. 

The point predictor for $Y$ is $\hat{Y} = \frac{1}{n}\sum_{i=1}^{n}Y_i$ (which is equal to $\hat{\mu}$), and the $1-\alpha$ *prediction interval* is given by

\begin{eqnarray}\label{pi}
[\underline{Y},\overline{Y}]  = \hat{Y} \pm t^{1-\alpha/2}_{n-1} \hat{\sigma\cdot }  \sqrt{1 + 1/n} 
\end{eqnarray}


The prediction interval in (\ref{pi}) can be obtained from the pivot

$$ T= \displaystyle \frac{ \hat{Y}  - Y}{\sigma \sqrt{1+1/n}} \cdot  \frac{\sigma }{\hat{\sigma}} \sim \frac{N(0,1)}{\sqrt{\chi^2_{n-1} /(n-1)}}\sim \mathcal{T}_{n-1}$$
and it has the property that

$$\mathrm{Pr}([\underline{Y},\overline{Y}]  \ni Y ) = 1-\alpha$$

The following simulation corroborates the coverage property of the predition interval:  

```{r}
sim = function(alpha=0.05, n=5, mu=0, sigma=1){
  y = rnorm(n, mean=mu, sd=sigma)
  haty = mean(y)
  k = qt(alpha/2, df = n-1, lower.tail = F)
  pi = haty + c(-1,1) * k * sqrt( var(y) * (1 + (1/n)) )
  y = rnorm(1, mean=mu, sd=sigma)
  cover = (y >= pi[1] & y <= pi[2])
  return(cover)
}
set.seed(123)
mean( replicate(100, sim(alpha=0.05) ))
```

## References

* Efron and Hastie (2016) Computer-Age Statistical Inference: Algorithms, Evidence, and Data Science, Cambridge University Press


