---
title: "Final projects evaluation"
description: |
  Statistical learning, attending students.

author:
  - name: Aldo Solari
    url: https://aldosolari.github.io/
    affiliation: DEMS
    affiliation_url: https://www.unimib.it/ateneo/dipartimenti-e-scuole/alla-scoperta-%E2%80%A6/dipartimento-economia-metodi-quantitativi-e-strategie-impresa
date: "`r Sys.Date()`"
output:
  radix::radix_article:
    toc: true
    toc_depth: 1
---

```{r startup, include = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = F, eval=T, message=F, warning=F, error=F, 
                      comment=NA, cache=F, R.options=list(width=220))
```

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(ggridges)

blog <- read_csv("SL_FP.csv")
names(blog)=c("Time","Team",
              "C1","C2","C3","C4",
              "S1","S2","P1","P2","P3","Score")
blog = blog %>% arrange(desc(Team))
```

# From the audience

### Communication

C1: Does the main point come through?

C2: Communication: Were the main ideas presented in an orderly and clear manner?

C3: Did the talk maintain the interest of the audience?

C4: Were the slides (or another instrument) appropriate and helpful to the audience?

```{r}
ggplot(blog, aes(x = C1+C2+C3+C4, y = Team, fill=factor(..quantile..))) +
  stat_density_ridges(
    geom = "density_ridges_gradient", calc_ecdf = TRUE,
    quantiles = 2, quantile_lines = TRUE
  ) + theme_bw() + guides(fill=FALSE)
```


### Statistics

S1: Statistical methods and/or critical thinking adequate?

S2:  How well do the findings support the conclusions?

```{r}
ggplot(blog, aes(x = S1 + S2, y = Team, fill=factor(..quantile..))) +
  stat_density_ridges(
    geom = "density_ridges_gradient", calc_ecdf = TRUE,
    quantiles = 2, quantile_lines = TRUE
  ) + theme_bw() + guides(fill=FALSE)
```


### Project

P1: States the objective, able to identify relevant questions

P2: Creativity

P3: Effort

```{r}
ggplot(blog, aes(x = P1 + P2 + P3, y = Team, fill=factor(..quantile..))) +
  stat_density_ridges(
    geom = "density_ridges_gradient", calc_ecdf = TRUE,
    quantiles = 2, quantile_lines = TRUE
  ) + theme_bw() + guides(fill=FALSE)
```


### Overall score

```{r}
ggplot(blog, aes(x = Score, y = Team, fill=factor(..quantile..))) +
  stat_density_ridges(
    geom = "density_ridges_gradient", calc_ecdf = TRUE,
    quantiles = 2, quantile_lines = TRUE
  ) + theme_bw() + guides(fill=FALSE) 
```

